The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2019-09-02 razzo-abstract.tex
Updates from Overleaf

% Two to three sentences of
% more detailed background, comprehensible to
% scientists in related disciplines.
Starting from heritable material and explicit assumptions,
Bayesian phylogenetics allows to infer a jointly-estimated 
phylogeny and parameter estimates distribution.
One of these assumptions is the speciation model, which
mathematically describes the branching process of a phylogeny in time.
Speciation models commonly assume that speciation events are independent and disjointed. Yet, such assumption may overlook the complexity of certain processes, where environmental changes promote speciation events in multiple lineages.
% Purpose of the paper
This new layer of complexity can be captured developing a novel ad hoc speciation model. However, the introduction of a new model could be not necessary if current models are capable of describing the process, under simpler assumptions.
Here we investigate the extent of the discrepancy produced by current BEAST2's tree priors, when trying to infer back trees where speciation can co-occur.

\iffalse
This research answers the question what the impact is of using a
species tree model that assumes speciation is independent, when it
is used on phylogenies created by a tree model in which speciation can
co-occur.
\fi

>>>>

% Two to three sentences of
% more detailed background, comprehensible to
% scientists in related disciplines.
Starting from heritable material and explicit assumptions,
Bayesian phylogenetics allows to infer a jointly-estimated 
phylogeny and parameter estimates distribution.
One of these assumptions is the speciation model, which
mathematically describes the branching process of a phylogeny in time.
Speciation models commonly assume that speciation events are independent and disjointed. Yet, such assumption may overlook the complexity of certain processes, where environmental changes promote speciation events in multiple lineages.
% Purpose of the paper
This new layer of complexity can be captured developing a novel ad hoc speciation model. However, the introduction of a new model could be not necessary if current models are already capable of describing the process to a satisfactory extent, under simpler assumptions.
Here we investigate the extent of the discrepancy produced by current BEAST2's tree priors, when trying to infer back trees where speciation can co-occur.

### 2019-09-02 razzo-methods.tex
Updates from Overleaf

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, to the common per-species speciation and extinction rates already present in the standard BD model. 
Additionally, MBD relies on two additional parameters, $\nu$ and $q$. 
The first, $\nu$, is the rate at which an environmental change is triggered.
When such event is triggered, each species present in the phylogeny at that moment has a probability $q$ to speciate at that time.
This kind of speciation is of a different nature respect to the one triggered by $\lambda$. In fact, whereas parameter $\lambda$ can be seen as describing a sympatric process, $\nu$ induces the rise of geographical barriers interrupting the gene flow \giovanni{@richel: please check if biology here is accurate} and leading to an allopatric speciation.
Even though multiple speciations can co-occur, polytomies are not allowed in such process as each species can speciate only once at the time.
A likelihood expression for the process is provided in \cite{mbd}.

@@

We simulate the speciation process in continuous time using the Doob-Gillespie algorithm, using the \verb;mbd_sim; function from the \verb;mbd; package (\citep{mbd}).
We let parameters vary using all possible combinations of values as shown in Table~\ref{tab:simulation_parameters}.
For each parameter setting, we generate $1000$ independent reconstructed trees of the same crown age.
\giovanni{Do we need to add more information on the Doob-Gillespie algorithm or it is overkilling?}

\input{razzo-tables/tab_simparameters.tex}

@@

From each MBD tree, we measure the impact of using the simpler BD tree prior on the inference, using the \verb;pirouette; R package, as described in detail in \citep{pirouette}.

In brief: \verb;pirouette; starts from a 'true' (but unobservable in nature) starting phylogeny, from which a DNA sequence alignment (which is observable in nature) is simulated. 
From each sequence alignment, a Bayesian inference is run, to obtain a posterior distribution of jointly-estimated trees and model parameter estimates.
By comparing the true tree and the posterior trees, an inference error distribution is generated. 
We use the twinning option available in \verb;pirouette; to measure a minimum and full error, in which the minimum error is caused by stochasticity in the full pipeline, where the full error is the added error from using an invalid but standard speciation model \giovanni{I am not sure I get this. What do you mean?}.

@@

For the experiments, we use both a hand-picked generative model
and a set of candidate models. We set a BD tree prior, JC69 site
model and strict clock model as the generative model. We used
all other combinations of four tree priors, two clock models and
five speciation models, resulting in a set of 39 candidate models.

>>>>

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, to the common per-species speciation and extinction rates already present in the standard BD model. 
Additionally, MBD relies on two additional parameters, $\nu$ and $q$. 
The first, $\nu$, is the rate at which an environmental change is triggered.
When such event is triggered, each species present in the phylogeny at that time has a probability $q$ to speciate.
This kind of speciation is of a different nature respect to the one triggered by $\lambda$. In fact, whereas parameter $\lambda$ can be seen as describing a sympatric process, $\nu$ induces the rise of geographical barriers interrupting the gene flow \giovanni{@richel: please check if biology here is accurate} and leading to an allopatric speciation.
Even though multiple speciations can co-occur, polytomies are not allowed in such process as each species can speciate only once at the time.
A likelihood expression for the process is provided in \cite{mbd}.

@@

We simulate the speciation process in continuous time exploiting the Doob-Gillespie algorithm, using the \verb;mbd_sim; function from the \verb;mbd; package (\citep{mbd}).
We let parameters vary using all possible combinations of values as shown in Table~\ref{tab:simulation_parameters}.
For each parameter setting, we generate $1000$ independent reconstructed trees with the same crown age.
\giovanni{Do we need to add more information on the Doob-Gillespie algorithm or it is overkilling?}

@@

From each MBD tree, we measure the impact of using the simpler BD tree prior on the inference, using the \verb;pirouette; R package, as described in detail in \citet{bilderbeek2019pirouette}.

In brief: \verb;pirouette; starts from a 'true' (but unobservable in nature) starting phylogeny, from which a DNA sequence alignment (which is instead observable) is simulated. 
From each sequence alignment, a Bayesian inference is run, to obtain a posterior distribution of jointly-estimated trees and model parameter estimates.
By comparing the true tree and the posterior trees, an inference error distribution is generated. 
We use the twinning option available in \verb;pirouette; to measure a minimum and full error, in which the minimum error is caused by stochasticity in the full pipeline, where the full error is the added error from using an invalid but standard speciation model \giovanni{I am not sure I get this. What do you mean?}.

@@

For the experiments, we use both a hand-picked generative model
and a set of candidate models. We set a BD tree prior, JC69 site
model and strict clock model as the generative model. We used
all other combinations of five tree priors, four site models and two clock models, resulting in a set of 39 candidate models.
